---
title: "Suicide Rates Data Analysis Project"
author: "Tanisha Damle"
format: 
  html:
    embed-resources: true
    code-tools: true
    code-summary: "Code"
---

# Introduction of Dataset

The dataset we're working with covers suicide rates and factors possibly linked to suicide from 1990 to 2022. It helps us understand how suicide relates to different aspects of our lives. The data includes information on suicide numbers, population, and economic conditions across regions and countries. We trust its reliability because it comes from respected sources like the World Health Organization Suicide Prevention and the World Bank Open Data.

# Questions Explored With This Dataset

Question 1: How do suicide rates vary across different age groups, genders, and countries, and how has this changed over time?

Question 2: Are there any trends in suicide rates that can be attributed to changes in access to mental health resources, as reflected in GNI per capita, and how does this vary across countries with different levels of economic development? 

Question 3: How does the employment population ratio and GDP per capita impact suicide rates, and is there a difference between genders and age groups?

---

# Visualizations and Analysis

## Introductory Data Analysis Plan

Before any visualization is created or any analysis is done for any question, the data needs to be cleaned up and a few libraries need to be loaded. We will need the "tidyverse" library in order to load the dataset and we will need "ggplot2" in order to visualize the data in a clean manner. Following this, we'll import the dataset into the R environment to enhance its readability and accessibility for subsequent tasks. Additionally, to streamline our visualization and analysis, we'll address any missing values within the dataset, ensuring a smooth and comprehensive analytical process. All of these steps are performed below: 

```{r, message=FALSE}
# load libraries
library(tidyverse)
library(ggplot2)
```

```{r, message=FALSE}
# Read the dataset and drop missing values
data <- read.csv("suicide_rates_1990-2022.csv")
data <- na.omit(data)
```

## Question 1

How do suicide rates vary across different age groups, genders, and countries, and how has this changed over time?

**Specific Data Analysis Plan**

For this question, we will first perform basic visualizations to display the data and view potential correlations. Data will be grouped by different groups within each sub-topic in the question, such as by age group, gender, and country. We will analyze correlations we see with a linear regression model done for each group and each sub-group by using the "lm" function. In order to do this, any outliers will be taken out as the linear regression model would be incorrect if the outliers were left in. The outliers will be calculated based on the sub-group average suicide rate's interquartile range. 

**Visualizations**

Age Group vs Year: 
```{r, message=FALSE}
# Group data by age group and year, and calculate average suicide rates
suicide_stats_by_age <- data %>%
  group_by(Year, AgeGroup) %>%
  summarise(AvgSuicideRate = mean(SuicideCount))

# Plotting trends over time for different age groups
ggplot(data = suicide_stats_by_age, aes(x = Year, y = AvgSuicideRate, color = AgeGroup)) +
  geom_point() +
  labs(title = "Trends in Suicide Rates Over Time by Age Group",
       x = "Year",
       y = "Average Suicide Rate",
       color = "Age Group")
```

Gender vs Year: 
```{r, message=FALSE}
# Group data by gender and year, and calculate average suicide rates
suicide_stats_by_gender <- data %>%
  group_by(Year, Sex) %>%
  summarize(AvgSuicideRate = mean(SuicideCount))

# Plotting trends over time for different genders
ggplot(data = suicide_stats_by_gender, aes(x = Year, y = AvgSuicideRate, color = Sex)) +
  geom_point() +
  labs(title = "Trends in Suicide Rates Over Time by Gender",
       x = "Year",
       y = "Average Suicide Rate",
       color = "Gender")
```

Region vs Year: 
```{r, message=FALSE}
# Group data by region and year, and calculate average suicide rates
suicide_stats_by_region <- data %>%
  group_by(Year, RegionName) %>%
  summarize(AvgSuicideRate = mean(SuicideCount))

# Plotting trends over time for different regions
ggplot(data = suicide_stats_by_region, aes(x = Year, y = AvgSuicideRate, color = RegionName)) +
  geom_point() +
  labs(title = "Trends in Suicide Rates Over Time by Region",
       x = "Year",
       y = "Average Suicide Rate",
       color = "Region")
```
**Outline Assumptions**

1. Linearity: In this analysis, we are're examining how suicide rates (dependent variable) change over time (independent variable) within each age group. The assumption of linearity suggests that the relationship between time (year) and suicide rates within each age group can be adequately represented by a straight line. With the visualizations, we can see that the observed patterns in this data align with this assumption because trends can be represented by a straight line.

2. Independence: This assumption implies that the errors (residuals) in predicting suicide rates for one year within an age group should not be related to errors from predicting suicide rates for another year within the same age group. Because each age group is separated by age and there is no overlap within individuals within one age group either, each observation's error is independent of others.

3. Normality: This assumption pertains to the distribution of errors around the regression line. Even though we are primarily interested in the relationship between time (year) and suicide rates, we can assume that the residuals from our regression models follow a normal distribution within each age group.

4. Equal Variance: We are investigating whether the variance in suicide rates remains constant over time within each age group. This assumption suggests that the spread of the residuals around the regression line should be consistent across different years within the same age group. We can assume that the variance of residuals is stable or exhibits patterns of increasing or decreasing variability over time.

Because these four assumptions can be made with this dataset, we can proceed and perform an inferential analysis with a linear regression. 

**Inferential Analysis**

Linear Model for Age Group vs Suicide Rate: 
```{r}

# Perform linear regression for the age group "0-14 years"
age_group_0_14 <- filter(suicide_stats_by_age, AgeGroup == "0-14 years")
lm_model_0_14 <- lm(AvgSuicideRate ~ Year, data = age_group_0_14)
residuals_0_14 <- residuals(lm_model_0_14)

Q1_0_14 <- quantile(residuals_0_14, 0.25)
Q3_0_14 <- quantile(residuals_0_14, 0.75)
IQR_0_14 <- Q3_0_14 - Q1_0_14

threshold_0_14 <- 1.5 * IQR_0_14
outliers_0_14 <- abs(residuals_0_14) > threshold_0_14
age_group_0_14 <- age_group_0_14[!outliers_0_14, ]

lm_model_0_14_new <- lm(AvgSuicideRate ~ Year, data = age_group_0_14)
coef(lm_model_0_14_new)


# Perform linear regression for the age group "15-24 years"
age_group_15_24 <- filter(suicide_stats_by_age, AgeGroup == "15-24 years")
lm_model_15_24 <- lm(AvgSuicideRate ~ Year, data = age_group_15_24)
residuals_15_24 <- residuals(lm_model_15_24)

Q1_15_24 <- quantile(residuals_15_24, 0.25)
Q3_15_24 <- quantile(residuals_15_24, 0.75)
IQR_15_24 <- Q3_15_24 - Q1_15_24

threshold_15_24 <- 1.5 * IQR_15_24
outliers_15_24 <- abs(residuals_15_24) > threshold_15_24
age_group_15_24 <- age_group_15_24[!outliers_15_24, ]

lm_model_15_24_new <- lm(AvgSuicideRate ~ Year, data = age_group_15_24)
coef(lm_model_15_24_new)

# Perform linear regression for the age group "25-34 years"
age_group_25_34 <- filter(suicide_stats_by_age, AgeGroup == "25-34 years")
lm_model_25_34 <- lm(AvgSuicideRate ~ Year, data = age_group_25_34)
residuals_25_34 <- residuals(lm_model_25_34)

Q1_25_34 <- quantile(residuals_25_34, 0.25)
Q3_25_34 <- quantile(residuals_25_34, 0.75)
IQR_25_34 <- Q3_25_34 - Q1_25_34

threshold_25_34 <- 1.5 * IQR_25_34
outliers_25_34 <- abs(residuals_25_34) > threshold_25_34
age_group_25_34 <- age_group_25_34[!outliers_25_34, ]

lm_model_25_34_new <- lm(AvgSuicideRate ~ Year, data = age_group_25_34)
coef(lm_model_25_34_new)

# Perform linear regression for the age group "35-54 years"
age_group_35_54 <- filter(suicide_stats_by_age, AgeGroup == "35-54 years")
lm_model_35_54 <- lm(AvgSuicideRate ~ Year, data = age_group_35_54)
residuals_35_54 <- residuals(lm_model_35_54)

Q1_35_54 <- quantile(residuals_35_54, 0.25)
Q3_35_54 <- quantile(residuals_35_54, 0.75)
IQR_35_54 <- Q3_35_54 - Q1_35_54

threshold_35_54 <- 1.5 * IQR_35_54
outliers_35_54 <- abs(residuals_35_54) > threshold_35_54
age_group_35_54 <- age_group_35_54[!outliers_35_54, ]

lm_model_35_54_new <- lm(AvgSuicideRate ~ Year, data = age_group_35_54)
coef(lm_model_35_54_new)

# Perform linear regression for the age group "55-74 years"
age_group_55_74 <- filter(suicide_stats_by_age, AgeGroup == "55-74 years")
lm_model_55_74 <- lm(AvgSuicideRate ~ Year, data = age_group_55_74)
residuals_55_74 <- residuals(lm_model_55_74)

Q1_55_74 <- quantile(residuals_55_74, 0.25)
Q3_55_74 <- quantile(residuals_55_74, 0.75)
IQR_55_74 <- Q3_55_74 - Q1_55_74

threshold_55_74 <- 1.5 * IQR_55_74
outliers_55_74 <- abs(residuals_55_74) > threshold_55_74
age_group_55_74 <- age_group_55_74[!outliers_55_74, ]

lm_model_55_74_new <- lm(AvgSuicideRate ~ Year, data = age_group_55_74)
coef(lm_model_55_74_new)

# Perform linear regression for the age group "75+ years"
age_group_75_plus <- filter(suicide_stats_by_age, AgeGroup == "75+ years")
lm_model_75_plus <- lm(AvgSuicideRate ~ Year, data = age_group_75_plus)
residuals_75_plus <- residuals(lm_model_75_plus)

Q1_75_plus <- quantile(residuals_75_plus, 0.25)
Q3_75_plus <- quantile(residuals_75_plus, 0.75)
IQR_75_plus <- Q3_75_plus - Q1_75_plus

threshold_75_plus <- 1.5 * IQR_75_plus
outliers_75_plus <- abs(residuals_75_plus) > threshold_75_plus
age_group_75_plus <- age_group_75_plus[!outliers_75_plus, ]

lm_model_75_plus_new <- lm(AvgSuicideRate ~ Year, data = age_group_75_plus)
coef(lm_model_75_plus_new)

```

Plotting Regression Line for Age Group vs Suicide Rate: 
```{r, message=FALSE}
# Create a data frame for coefficients of each age group
coefficients_df <- data.frame(
  AgeGroup = c("0-14 years", "15-24 years", "25-34 years", "35-54 years", "55-74 years", "75+ years"),
  Intercept = c(coef(lm_model_0_14_new)[1], coef(lm_model_15_24_new)[1], coef(lm_model_25_34_new)[1],
                coef(lm_model_35_54_new)[1], coef(lm_model_55_74_new)[1], coef(lm_model_75_plus_new)[1]),
  Year = c(coef(lm_model_0_14_new)[2], coef(lm_model_15_24_new)[2], coef(lm_model_25_34_new)[2],
           coef(lm_model_35_54_new)[2], coef(lm_model_55_74_new)[2], coef(lm_model_75_plus_new)[2]),
  Order = c(1, 2, 3, 4, 5, 6) # Add an order column
)

# Sort coefficients_df by color
coefficients_df <- coefficients_df[order(coefficients_df$Order), ]

# Plotting regression lines for each age group
ggplot() +
  geom_smooth(data = age_group_0_14, aes(x = Year, y = AvgSuicideRate, color = "0-14 years"), 
              method = "lm", se = FALSE) +
  geom_smooth(data = age_group_15_24, aes(x = Year, y = AvgSuicideRate, color = "15-24 years"), 
              method = "lm", se = FALSE) +
  geom_smooth(data = age_group_25_34, aes(x = Year, y = AvgSuicideRate, color = "25-34 years"), 
              method = "lm", se = FALSE) +
  geom_smooth(data = age_group_35_54, aes(x = Year, y = AvgSuicideRate, color = "35-54 years"), 
              method = "lm", se = FALSE) +
  geom_smooth(data = age_group_55_74, aes(x = Year, y = AvgSuicideRate, color = "55-74 years"), 
              method = "lm", se = FALSE) +
  geom_smooth(data = age_group_75_plus, aes(x = Year, y = AvgSuicideRate, color = "75+ years"), 
              method = "lm", se = FALSE) +
  labs(title = "Regression Lines for Suicide Rates by Age Group",
       x = "Year",
       y = "Average Suicide Rate",
       color = "Age Group") +
  scale_color_viridis_d() +
  guides(color = guide_legend(title = "Age Group"))

```

Linear Model for Gender vs Suicide Rate:

```{r}
# Perform linear regression for the gender group "Female"
female_data <- filter(suicide_stats_by_gender, Sex == "Female")
lm_female <- lm(AvgSuicideRate ~ Year, data = female_data)
residuals_female <- residuals(lm_female)

Q1_female <- quantile(residuals_female, 0.25)
Q3_female <- quantile(residuals_female, 0.75)
IQR_female <- Q3_female - Q1_female

threshold_female <- 1.5 * IQR_female
outliers_female <- abs(residuals_female) > threshold_female
female_data <- female_data[!outliers_female, ]

lm_female_new <- lm(AvgSuicideRate ~ Year, data = female_data)
coef(lm_female_new)

# Perform linear regression for the gender group "Male"
male_data <- filter(suicide_stats_by_gender, Sex == "Male")
lm_male <- lm(AvgSuicideRate ~ Year, data = male_data)
residuals_male <- residuals(lm_male)

Q1_male <- quantile(residuals_male, 0.25)
Q3_male <- quantile(residuals_male, 0.75)
IQR_male <- Q3_male - Q1_male

threshold_male <- 1.5 * IQR_male
outliers_male <- abs(residuals_male) > threshold_male
male_data <- male_data[!outliers_male, ]

lm_male_new <- lm(AvgSuicideRate ~ Year, data = male_data)
coef(lm_male_new)

```


Plotting Regression Line for Gender vs Suicide Rate: 

```{r, message=FALSE}
# Create a data frame for coefficients of each gender group
coefficients_gender_df <- data.frame(
  Sex = c("Female", "Male"),
  Intercept = c(coef(lm_female_new)[1], coef(lm_male_new)[1]),
  Year = c(coef(lm_female_new)[2], coef(lm_male_new)[2]),
  Order = c(1, 2) # Add an order column
)

# Sort coefficients_gender_df by color
coefficients_gender_df <- coefficients_gender_df[order(coefficients_gender_df$Order), ]

# Plotting regression lines for each gender group
ggplot() +
  geom_smooth(data = female_data, aes(x = Year, y = AvgSuicideRate, color = "Female"), 
              method = "lm", se = FALSE) +
  geom_smooth(data = male_data, aes(x = Year, y = AvgSuicideRate, color = "Male"), 
              method = "lm", se = FALSE) +
  labs(title = "Regression Lines for Suicide Rates by Gender",
       x = "Year",
       y = "Average Suicide Rate",
       color = "Gender") +
  scale_color_viridis_d() +
  guides(color = guide_legend(title = "Gender"))

```

Linear Model for Country vs Suicide Rate:

```{r}
# Filter data for each country group and perform linear regression
# Africa
africa_data <- filter(suicide_stats_by_region, RegionName == "Africa")
lm_africa <- lm(AvgSuicideRate ~ Year, data = africa_data)
residuals_africa <- residuals(lm_africa)
Q1_africa <- quantile(residuals_africa, 0.25)
Q3_africa <- quantile(residuals_africa, 0.75)
IQR_africa <- Q3_africa - Q1_africa
threshold_africa <- 1.5 * IQR_africa
outliers_africa <- abs(residuals_africa) > threshold_africa
africa_data <- africa_data[!outliers_africa, ]
lm_africa_new <- lm(AvgSuicideRate ~ Year, data = africa_data)
coef(lm_africa_new)

# Asia
asia_data <- filter(suicide_stats_by_region, RegionName == "Asia")
lm_asia <- lm(AvgSuicideRate ~ Year, data = asia_data)
residuals_asia <- residuals(lm_asia)
Q1_asia <- quantile(residuals_asia, 0.25)
Q3_asia <- quantile(residuals_asia, 0.75)
IQR_asia <- Q3_asia - Q1_asia
threshold_asia <- 1.5 * IQR_asia
outliers_asia <- abs(residuals_asia) > threshold_asia
asia_data <- asia_data[!outliers_asia, ]
lm_asia_new <- lm(AvgSuicideRate ~ Year, data = asia_data)
coef(lm_asia_new)

# Central and South America
central_south_america_data <- filter(suicide_stats_by_region, RegionName == "Central and South America")
lm_central_south_america <- lm(AvgSuicideRate ~ Year, data = central_south_america_data)
residuals_central_south_america <- residuals(lm_central_south_america)
Q1_central_south_america <- quantile(residuals_central_south_america, 0.25)
Q3_central_south_america <- quantile(residuals_central_south_america, 0.75)
IQR_central_south_america <- Q3_central_south_america - Q1_central_south_america
threshold_central_south_america <- 1.5 * IQR_central_south_america
outliers_central_south_america <- abs(residuals_central_south_america) > threshold_central_south_america
central_south_america_data <- central_south_america_data[!outliers_central_south_america, ]
lm_central_south_america_new <- lm(AvgSuicideRate ~ Year, data = central_south_america_data)
coef(lm_central_south_america_new)

# Europe
europe_data <- filter(suicide_stats_by_region, RegionName == "Europe")
lm_europe <- lm(AvgSuicideRate ~ Year, data = europe_data)
residuals_europe <- residuals(lm_europe)
Q1_europe <- quantile(residuals_europe, 0.25)
Q3_europe <- quantile(residuals_europe, 0.75)
IQR_europe <- Q3_europe - Q1_europe
threshold_europe <- 1.5 * IQR_europe
outliers_europe <- abs(residuals_europe) > threshold_europe
europe_data <- europe_data[!outliers_europe, ]
lm_europe_new <- lm(AvgSuicideRate ~ Year, data = europe_data)
coef(lm_europe_new)

# Oceania
oceania_data <- filter(suicide_stats_by_region, RegionName == "Oceania")
lm_oceania <- lm(AvgSuicideRate ~ Year, data = oceania_data)
residuals_oceania <- residuals(lm_oceania)
Q1_oceania <- quantile(residuals_oceania, 0.25)
Q3_oceania <- quantile(residuals_oceania, 0.75)
IQR_oceania <- Q3_oceania - Q1_oceania
threshold_oceania <- 1.5 * IQR_oceania
outliers_oceania <- abs(residuals_oceania) > threshold_oceania
oceania_data <- oceania_data[!outliers_oceania, ]
lm_oceania_new <- lm(AvgSuicideRate ~ Year, data = oceania_data)
coef(lm_oceania_new)

# North America and the Caribbean
north_america_caribbean_data <- filter(suicide_stats_by_region, RegionName == "North America and the Caribbean")
lm_north_america_caribbean <- lm(AvgSuicideRate ~ Year, data = north_america_caribbean_data)
residuals_north_america_caribbean <- residuals(lm_north_america_caribbean)
Q1_north_america_caribbean <- quantile(residuals_north_america_caribbean, 0.25)
Q3_north_america_caribbean <- quantile(residuals_north_america_caribbean, 0.75)
IQR_north_america_caribbean <- Q3_north_america_caribbean - Q1_north_america_caribbean
threshold_north_america_caribbean <- 1.5 * IQR_north_america_caribbean
outliers_north_america_caribbean <- abs(residuals_north_america_caribbean) > threshold_north_america_caribbean
north_america_caribbean_data <- north_america_caribbean_data[!outliers_north_america_caribbean, ]
lm_north_america_caribbean_new <- lm(AvgSuicideRate ~ Year, data = north_america_caribbean_data)
coef(lm_north_america_caribbean_new)

```

Plotting Regression Line for Country vs Suicide Rate: 

```{r, message=FALSE}

# Create coefficient dataframes for each country group
coefficients_country_df <- data.frame(
  Country = c("Africa", "Asia", "Central and South America", "Europe", "Oceania", "North America and the Caribbean"),
  Intercept = c(coef(lm_africa_new)[1], coef(lm_asia_new)[1], coef(lm_central_south_america_new)[1],
                coef(lm_europe_new)[1], coef(lm_oceania_new)[1], coef(lm_north_america_caribbean_new)[1]),
  Year = c(coef(lm_africa_new)[2], coef(lm_asia_new)[2], coef(lm_central_south_america_new)[2],
           coef(lm_europe_new)[2], coef(lm_oceania_new)[2], coef(lm_north_america_caribbean_new)[2]),
  Order = c(1, 2, 3, 4, 5, 6) # Add an order column
)

# Sort coefficients_country_df by order
coefficients_country_df <- coefficients_country_df[order(coefficients_country_df$Order), ]

# Plotting regression lines for each country group
ggplot() +
  geom_smooth(data = africa_data, aes(x = Year, y = AvgSuicideRate, color = "Africa"), 
              method = "lm", se = FALSE) +
  geom_smooth(data = asia_data, aes(x = Year, y = AvgSuicideRate, color = "Asia"), 
              method = "lm", se = FALSE) +
  geom_smooth(data = central_south_america_data, aes(x = Year, y = AvgSuicideRate, color = "Central and South America"), 
              method = "lm", se = FALSE) +
  geom_smooth(data = europe_data, aes(x = Year, y = AvgSuicideRate, color = "Europe"), 
              method = "lm", se = FALSE) +
  geom_smooth(data = oceania_data, aes(x = Year, y = AvgSuicideRate, color = "Oceania"), 
              method = "lm", se = FALSE) +
  geom_smooth(data = north_america_caribbean_data, aes(x = Year, y = AvgSuicideRate, color = "North America and the Caribbean"), 
              method = "lm", se = FALSE) +
  labs(title = "Regression Lines for Suicide Rates by Country",
       x = "Year",
       y = "Average Suicide Rate",
       color = "Country") +
  scale_color_viridis_d() +
  guides(color = guide_legend(title = "Country"))

```


